Bayesian Non-parametric Inference for Multivariate\\ Peaks-over-threshold Models

### 2024-02-01 methodology_projgamma.tex
changes

To obtain a distribution on 
${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ 
we start with a vector in 
$\bm{x} \in {\mathbb R}^d_+$, and normalize it to obtain 
$\bm{y} = \bm{x}/\|\bm{x}\|_p \in {\mathbb S}_p^{d-1}$. 
A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent
  univariate Gamma distributions. Let
    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. 
    For any finite $p>0$, letting 
   $ y_d = (1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p)^{1/p}$,
    the transformation
  \begin{equation}
    \label{eqn:pnormt}
    \begin{aligned}
    T(x_1,\ldots,x_d) &= \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) \\
                        &= (r,y_1,\ldots,y_{d-1})
    \end{aligned}
  \end{equation}
  is invertible with
  \begin{equation}
    \label{eqn:pnormtinv}
    \begin{aligned}
    &\hspace{-0.5cm}T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)\\
      & = \left(ry_1,\ldots,ry_{d-1}, 
        r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
    \end{aligned}
  \end{equation}
  The Jacobian of the transformation takes the form
  \begin{equation}
    \label{eqn:pnormjac}
    \begin{aligned}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right. & \\
        &\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p
          \left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
    \end{aligned}
  \end{equation}
  The normalization provided by $T$ maps a vector in ${\mathbb R}_+^d$ onto
  ${\mathbb S}_p^{d-1} $. With a slight abuse of terminology we refer to it as a
  projection. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
  \begin{equation}
  \label{joint}
    \begin{aligned}
    f(r,\bm{ y}) = &\prod_{\ell = 1}^{d}
      \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right] \\
      &\hspace{0.5cm}\times r^{d-1}\left[y_d +
            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right].
    \end{aligned}
  \end{equation}
  Integrating out $r$ yields the resulting \emph{Projected Gamma} density
  \begin{equation}
    \label{eqn:projgamma}
    \begin{aligned}
    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) &=
          \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
                y_{\ell}^{\alpha_{\ell} - 1}\right]\\
      &\hspace{0.5cm}\times \left[y_d +
          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right]\\
      &\hspace{0.5cm}\times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}},
    \end{aligned}
  \end{equation}
  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any finite $p>0$. To avoid identifiability 
  problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
  \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
  for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.  
  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in
  Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.

@@

Unfortunately, in the limit when $p\rightarrow \infty$, the normalizing 
  transformation is not differentiable. Thus, a closed form expression 
  like Equation (\ref{eqn:projgamma}) for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
  available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach
  $\mathbb{S}_{\infty}^{d-1}$.  With that in mind, our strategy consists of describing the angular
  distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the 
  transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the 
  subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and 
  project the observations onto ${\mathbb S}_p^{d-1}$; 
  (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting 
  data and obtain samples from the fitted model; 
  (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
  For step (iv) we use a Bayesian approach that is implemented using a purposely 
  developed Markov chain Monte Carlo sampler described in the next section.

@@

Equation \ref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible 
  under our proposed model. For the analysis of extreme values it is of interest to calculate the multivariate 
  conditional survival function. The following result provides the relevant expression, as a function of the 
  angular measure.
  \begin{prop}
  Assume the same conditions of Proposition \ref{ppchi}.   Let $\alpha \subset \{1, \ldots ,d\}$ be a 
  collections of indexes.  Then     
  \begin{equation} \label{eqn:condsurv}
    \begin{aligned}
    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} 
        Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
    \end{aligned}
  \end{equation}
  \end{prop}  
  The proof uses a similar approach to the proof of Proposition \ref{ppchi}.

Equations \ref{eqn:chi_ij} and \ref{eqn:condsurv} provide relevant tools for inference on the tail 
  behavior of the joint distribution of the observations. The expressions can be readily calculated 
  within a sample-based inferential approach like the one considered in the following section.

\subsubsection{Inference for the projected Gamma mixture model}
In a sample-based inference approach, for a given iteration the Dirichlet process mixture model 
    groups observations into stochastically assigned clusters, where members of a cluster share 
    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for Equation~(\ref{eqn:dppg}), 
    let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
    the number of extant clusters, not including any singleton containing observation $i$. Under this model, 
    the probability of cluster membership for a given observation is proportional to
\begin{equation*}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm} \\ % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j),%\hspace{0.5cm} % &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
where the top case is iterating over extant clusters $j = 1,\ldots, J^{-i}$, and the bottom case is
    for a \emph{new} cluster. If $G_0$ is not a conjugate prior for the kernel density, the integral in the 
    above formula may not be available in closed form. We sidestep this using Algorithm 8 from \cite{neal2000}: 
    by Monte Carlo integration, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for
    $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, we sample the cluster indicator 
    $\gamma_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
\begin{equation}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm}  % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right). % \hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
Again, the top case is iterating over extant clusters, and now the bottom case is iterating 
    over new \emph{candidate} clusters.  If a candidate cluster is selected,
%{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
    then $\gamma_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.

@@

for each observation. Using Equation \eqref{joint} we observe that the 

@@

@@

Note that the ordering of the products can be reversed in the likelihood, indicating that given $\bm{r}$, 
the likelihood becomes separable by dimension.

@@

This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.  An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form

@@

model, we use the full conditional in Equation~(\ref{eqn:alpha1update}) for all $\ell$, and omit 

>>>>

To obtain a distribution on 
${\mathbb S}_{p}^{d-1}=\{\bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \| \bm{s}\|_{p} = 1\}$ 
we start with a vector in 
$\bm{x} \in {\mathbb R}^d_+$, and normalize it to obtain 
$\bm{y} = \bm{x}/\|\bm{x}\|_p \in {\mathbb S}_p^{d-1}$. 
A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent
  univariate Gamma distributions. Let
    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. 
    For any finite $p>0$, letting 
   $ y_d = (1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p)^{1/p}$,
    the transformation
  \begin{equation}
    \label{eqn:pnormt}
    \begin{aligned}
    T(x_1,\ldots,x_d) &= \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) \\
                        &= (r,y_1,\ldots,y_{d-1})
    \end{aligned}
  \end{equation}
  is invertible with
  \begin{equation}
    \label{eqn:pnormtinv}
    \begin{aligned}
    &\hspace{-0.5cm}T^{-1}\left(r,y_1,\ldots,y_{d-1}\right)\\
      & = \left(ry_1,\ldots,ry_{d-1}, 
        r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
    \end{aligned}
  \end{equation}
  The Jacobian of the transformation takes the form
  \begin{equation}
    \label{eqn:pnormjac}
    \begin{aligned}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} \right. & \\
        &\hspace{-2cm}+ \left. {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p
          \left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
    \end{aligned}
  \end{equation}
  The normalization provided by $T$ maps a vector in ${\mathbb R}_+^d$ onto
  ${\mathbb S}_p^{d-1} $. With a slight abuse of terminology we refer to it as a
  projection. Using Equations \eqref{eqn:pnormt}--\eqref{eqn:pnormjac} we have the joint density
  \begin{equation}
  \label{joint}
    \begin{aligned}
    f(r,\bm{ y}) = &\prod_{\ell = 1}^{d}
      \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right] \\
      &\hspace{0.5cm}\times r^{d-1}\left[y_d +
            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right].
    \end{aligned}
  \end{equation}
  Integrating out $r$ yields the resulting \emph{Projected Gamma} density
  \begin{equation}
    \label{eqn:projgamma}
    \begin{aligned}
    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) &=
          \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
                y_{\ell}^{\alpha_{\ell} - 1}\right]\\
      &\hspace{0.5cm}\times \left[y_d +
          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p y_d^{1 - p}\right]\\
      &\hspace{0.5cm}\times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}},
    \end{aligned}
  \end{equation}
  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any finite $p>0$. To avoid identifiability 
  problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
  \cite{nunez2019} obtain the density in Equation~\eqref{eqn:projgamma}
  for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.  
  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in
  Equation~\eqref{eqn:projgamma} corresponds to that of a Dirichlet distribution.

@@

Unfortunately, in the limit when $p\rightarrow \infty$, the normalizing 
  transformation is not differentiable. Thus, a closed form expression 
  like Equation~\eqref{eqn:projgamma} for the projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
  available. Instead, we observe that for a sufficiently large $p$, $\mathbb{S}_p^{d-1}$ will approach
  $\mathbb{S}_{\infty}^{d-1}$.  With that in mind, our strategy consists of describing the angular
  distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the 
  transformation in Equation~\eqref{eqn:standardization} to the original data; (ii) Obtain the 
  subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and 
  project the observations onto ${\mathbb S}_p^{d-1}$; 
  (iv) Fit the model in Equation~\eqref{eqn:PGmix} to the resulting 
  data and obtain samples from the fitted model; 
  (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
  For step (iv) we use a Bayesian approach that is implemented using a purposely 
  developed Markov chain Monte Carlo sampler described in the next section.

@@

Equation~\eqref{eqn:chi_ij} implies that $\chi_{\jmath\ell}>0$, and so, no asymptotic independence is possible 
  under our proposed model. For the analysis of extreme values it is of interest to calculate the multivariate 
  conditional survival function. The following result provides the relevant expression, as a function of the 
  angular measure.
  \begin{prop}
  Assume the same conditions of Proposition \ref{ppchi}.   Let $\alpha \subset \{1, \ldots ,d\}$ be a 
  collections of indexes.  Then     
  \begin{equation} \label{eqn:condsurv}
    \begin{aligned}
    &\hspace{-0.5cm}\mathrm{Pr}\left[\bigcap_{\ell \in \alpha} 
        Z_\ell > z_\ell \mid \bigcap_{\ell\not\in\alpha} Z_\ell > z_\ell\right] \\
    &= \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
    \end{aligned}
  \end{equation}
  \end{prop}  
  The proof uses a similar approach to the proof of Proposition \ref{ppchi}.

Equations~\eqref{eqn:chi_ij} and \eqref{eqn:condsurv} provide relevant tools for inference on the tail 
  behavior of the joint distribution of the observations. The expressions can be readily calculated 
  within a sample-based inferential approach like the one considered in the following section.

\subsubsection{Inference for the projected Gamma mixture model}
In a sample-based inference approach, for a given iteration the Dirichlet process mixture model 
    groups observations into stochastically assigned clusters, where members of a cluster share 
    distributional parameters \citep{muller2015,ascolani2022}. Building out the methods of inference for Equation~\eqref{eqn:dppg}, 
    let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
    the number of extant clusters, not including any singleton containing observation $i$. Under this model, 
    the probability of cluster membership for a given observation is proportional to
\begin{equation*}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm} \\ % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j),%\hspace{0.5cm} % &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
where the top case is iterating over extant clusters $j = 1,\ldots, J^{-i}$, and the bottom case is
    for a \emph{new} cluster. If $G_0$ is not a conjugate prior for the kernel density, the integral in the 
    above formula may not be available in closed form. We sidestep this using Algorithm 8 from \cite{neal2000}: 
    by Monte Carlo integration, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for
    $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, we sample the cluster indicator 
    $\gamma_i$ from extant or candidate clusters, where the probability of cluster membership is proportional to
\begin{equation}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\\ % \hspace{0.5cm}  % &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right). % \hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
Again, the top case is iterating over extant clusters, and now the bottom case is iterating 
    over new \emph{candidate} clusters.  If a candidate cluster is selected,
%{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
    then $\gamma_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.

@@

for each observation. Using Equation~\eqref{joint} we observe that the 

@@

    \label{eqn:alphabetafc}

@@

Note that the ordering of the products can be reversed in Equation~\eqref{eqn:alphabetafc}, 
indicating that with appropriate choice of centering distribution, the full conditionals 
for $\bm{\alpha}_j,\bm{\beta}_j$ become separable by dimension.

@@

This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  
We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  
We refer to this model as the \emph{projected gamma--gamma} (PG--G) model.  An advantage of the PG--G model 
is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  
Then, the full conditional for $\alpha_{j\ell}$ takes the form

@@

model, we use the full conditional in Equation~\eqref{eqn:alpha1update} for all $\ell$, and omit 
